Understanding Experimental Data in the Rational Speech Act Framework

### 2018-02-14 models/linking hypotheses/linkinghypotheses.rmd
made edits to linkinghypotheses file

---
title: "Linking Hypotheses and SI-Paradigms"
author: "Brandon Waldon"
output: github_document
---

@@

@@

##Setting up the RSA model

In the RSA model, the function S1 returns the probability of an utterance given the world state the speaker intends to communicate.

@@

Let's first explore a linking hypothesis that makes the wrong predictions in this regard. We might hypothesize that speakers, when confronted with a world state w (i.e. the configuration of animals on the card in a given SI-Paradigms trial) and an utterance u (the guess in that trial), have access to the probability that the S1 distribution assigns to u in w.

@@

On this simplistic straw-man story, we might say that all that happens next is that speakers sample from a Bernoulli distribution where the probability of success is equal to S1(u|w). If that sample is a success, the utterance is rated as Right; if failure, the utterance is rated as Wrong.

@@

P(Participant answers "Right"|u,w) = S1(u|w)

Perhaps the straw-man hypothesis is basically correct in conceptualizing the categorical response as a probabilistic process. If we accept the basic story that the categorical response can be represented as P("Right"|u,w), then we need a way to link P("Right"|u,w) to our RSA speaker distribution S1(u|w) - in a way that gets better empirical coverage than the straw-man hypothesis explored above.

One alternative would be to say that P("Right"|u,w) is a function of, but not identical to, the probability of utterance u given world state w in the S1 model. The challenge then becomes to specify this function. Logistic regression provides one possible solution. This type of regression returns a model of the log odds of some binary categorical response (e.g. "Right"/"Wrong") as a function of an independent variable (e.g. speaker probability of utterance given a world state).

Exploring this requires some transformation to our results. We continue to look at the subset of the SI-paradigms pilot data where response was binary "Right"/"Wrong". Now, for every row in the data (each row corresponds to one response), we add a value S1(u|w) in a new column (speaker_probability) of the data frame, where S1(u|w) is the speaker probability of u (the 'guess' presented on the screen) given w (the configuration of animals on the card).

@@

Another conceivable hypothesis is that for some utterance u in world state w, u is "Right" so long as S1(u|w) exceeds some probability threshold $\theta$.

We can enrich the pragmatic speaker model provided above to model such a function. First, we specify a threshold function which takes as its input a world state w and a probability threshold $\theta$ and returns only those utterances for which S1(u|w) > $\theta$ is true. The following model also specifies a cost function - for now, we'll keep the cost uniform on utterances.

@@

For the experimental conditions in which more than two choices were provided, we need to specify an enriched linking function. For the ternary choice condition, we need a function that takes as its input an utterance, a world state, and two thresholds $\theta_1$ and $\theta_2$. An utterance is "Wrong" in world state w just in case S1(u|w) is less than $\theta_1$. It is "Right" just in case S1(u|w) > $\theta_2$. It is "Neither" just in case  $\theta_1$ < S1(u|w) < $\theta_2$.

@@

Yet another possibility is that for some utterance u in world state w, P("Right"|u,w) is a function of S1'(u|w), where S1' is a transformation of the pragmatic speaker distribution. More specifically, S1' is a re-normalized version of S1 after low-probability utterances are filtered from S1 (according to some threshold of probability).

In order to explore this hypothesis, we first need to model S1' (specified below as speaker_filtered). Speaker_filtered is a function of utterancePrior_filtered - an utterance prior function which takes as its input only high-probability utterances; that is, utterances for which S1(u|w) > $\theta$ is true.

@@

Let's explore this new S1' distribution over utterances, given a world state ['X','Y'] and that we are only interested in those utterances for which it is true that S1(u|w) > 0.12.

@@

With this model, we could revisit the logistic regression linking function explored above: the log odds of categorical response is a function of the (filtered) speaker probability of that utterance in a particular world state w and given a probability threshold $\theta$.

>>>>

---
title: "Understanding Experimental Data in the Rational Speech Act Framework"
author: "Brandon Waldon"
output: github_document
---

@@

## Background

Two contributions: one empirical, one theoretical

**'Sentence verification' paradigms and scalar implicature**

- Lots of things can affect speakers' judgments about scalar implicature.

- Pragmatic judgments change according to ``dependent measure used to probe linguistic intuitions" - e.g. 'sentence verification' tasks, 'word probability rating' tasks, and 'sentence interpretation' tasks (Degen and Goodman 2014: 1). Moreover, Degen and Goodman (2014) show that some experimental paradigms are more sensitive to contextual manipulations than others.

- Building off that work, we provide evidence that judgments in 'sentence verification' tasks are furthermore sensitive to the number of response options participants have when completing the task.

**Understanding what is measured in 'sentence verification' paradigms**

- The typical neo-Gricean way of understanding participant behavior on truth value judgment task experiments (to the extent linking hypotheses are discussed in that literature): a participant either does or does not `calculate' an implicature when presented some sentence S. The evaluation of the sentence S as "Right"/"Wrong", "True"/"False", "Correct"/"Incorrect" depends on whether an implicature was derived and whether or not that pragmatically enriched interpretation is actually used to verify S.

- Another proposal: building off Degen and Goodman (2014), participants' behavior on these tasks can be modeled directly as a function of a Rational Speech Act model of a pragmatic speaker.

@@

## Setting up the RSA model

In the RSA model, the function $S_1$ returns the probability of an utterance given the world state the speaker intends to communicate.

@@

Let's first explore a linking hypothesis that makes the wrong predictions in this regard. We might hypothesize that speakers, when confronted with a world state w (i.e. the configuration of animals on the card in a given SI-Paradigms trial) and an utterance u (the guess in that trial), have access to the probability that the $S_1$ distribution assigns to u in w.

@@

On this simplistic straw-man story, we might say that all that happens next is that speakers sample from a Bernoulli distribution where the probability of success is equal to $S_1$(u|w). If that sample is a success, the utterance is rated as Right; if failure, the utterance is rated as Wrong.

@@

P(Participant answers "Right"|u,w) = $S_1$(u|w)

Perhaps the straw-man hypothesis is basically correct in conceptualizing the categorical response as a probabilistic process. If we accept the basic story that the categorical response can be represented as P("Right"|u,w), then we need a way to link P("Right"|u,w) to our RSA speaker distribution $S_1$(u|w) - in a way that gets better empirical coverage than the straw-man hypothesis explored above.

One alternative would be to say that P("Right"|u,w) is a function of, but not identical to, the probability of utterance u given world state w in the $S_1$ model. The challenge then becomes to specify this function. Logistic regression provides one possible solution. This type of regression returns a model of the log odds of some binary categorical response (e.g. "Right"/"Wrong") as a function of an independent variable (e.g. speaker probability of utterance given a world state).

Exploring this requires some transformation to our results. We continue to look at the subset of the SI-paradigms pilot data where response was binary "Right"/"Wrong". Now, for every row in the data (each row corresponds to one response), we add a value $S_1$(u|w) in a new column (speaker_probability) of the data frame, where $S_1$(u|w) is the speaker probability of u (the 'guess' presented on the screen) given w (the configuration of animals on the card).

@@

Another conceivable hypothesis is that for some utterance u in world state w, u is "Right" so long as $S_1$(u|w) exceeds some probability threshold $\theta$.

We can enrich the pragmatic speaker model provided above to model such a function. First, we specify a threshold function which takes as its input a world state w and a probability threshold $\theta$ and returns only those utterances for which $S_1$(u|w) > $\theta$ is true. The following model also specifies a cost function - for now, we'll keep the cost uniform on utterances.

@@

For the experimental conditions in which more than two choices were provided, we need to specify an enriched linking function. For the ternary choice condition, we need a function that takes as its input an utterance, a world state, and two thresholds $\theta_1$ and $\theta_2$. An utterance is "Wrong" in world state w just in case $S_1$(u|w) is less than $\theta_1$. It is "Right" just in case $S_1$(u|w) > $\theta_2$. It is "Neither" just in case  $\theta_1$ < $S_1$(u|w) < $\theta_2$.

@@

Yet another possibility is that for some utterance u in world state w, P("Right"|u,w) is a function of $S_1$'(u|w), where $S_1$' is a transformation of the pragmatic speaker distribution. More specifically, $S_1$' is a re-normalized version of $S_1$ after low-probability utterances are filtered from $S_1$ (according to some threshold of probability).

In order to explore this hypothesis, we first need to model $S_1$' (specified below as speaker_filtered). Speaker_filtered is a function of utterancePrior_filtered - an utterance prior function which takes as its input only high-probability utterances; that is, utterances for which $S_1$(u|w) > $\theta$ is true.

@@

Let's explore this new $S_1$' distribution over utterances, given a world state ['X','Y'] and that we are only interested in those utterances for which it is true that $S_1$(u|w) > 0.12.

@@

With this model, we could revisit the logistic regression linking function explored above: the log odds of categorical response is a function of the (filtered) speaker probability of that utterance in a particular world state w and given a probability threshold $\theta$ (in the example below, $\theta$ is 0.12 - a totally arbitrary value).
